### 2016-05-13 text/Abstract.tex
Add interactive essay to abstract

%!TEX root = ../main.tex
\pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer

@@

\chapter*{Abstract} % Abstract name
The development of programming languages needs to reflect important changes in the way
programs execute. In recent years, this has included \eg~the development of parallel programming 
models (in reaction to the multi-core revolution) or improvements in data access technologies. 
This thesis is a response to another such revolution -- the diversification of devices and 
systems where programs run.

@@

We identify a number of program properties that were not connected before, but model some notion 
of context. Our examples include tracking different execution platforms (and their versions) 
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as 
variable usage (\eg~in liveness analaysis and linear logics) or past values in 
stream-based dataflow programming.

Our first contribution is the discovery of the connection between the above examples and
their novel presentation in the form of calculi (\emph{coeffect systems}). The presented type 
systems and formal semantics highlight the relationship between different notions of context. 
Our second and third contributions are two unified coeffect calculi that capture commonalities 
in the presented examples. In particular, our \emph{flat coeffect calculus} models languages 
with contextual properties of the execution environment and our \emph{structural coeffect 
calculus} models languages where the contextual properties are attached to the variable usage.
We close the thesis with a \emph{unified coeffect calculus} that captures the two notions
using a single parameterized system.

Although the focus of this thesis is on the syntactic properties of the presented 
systems, we also discuss their category-theoretical motivation. We introduce the notion of
an \emph{indexed} comonad (based on dualisation of the well-known monad structure) and show 
how they provide semantics of the two coeffect calculi.

>>>>

\pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer

@@

\chapter*{Abstract} % Abstract name
The development of programming languages needs to reflect important changes in the way
programs execute. In recent years, this has included the development of parallel programming
models (in reaction to the multi-core revolution) or improvements in data access technologies.
This thesis is a response to another such revolution -- the diversification of devices and
systems where programs run.

@@

We identify a number of program properties that were not connected before, but model some notion
of context. Our examples include tracking different execution platforms (and their versions)
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as
variable usage (\eg~in liveness analaysis and linear logics) or past values in
stream-based dataflow programming. Our first contribution is the discovery of the connection
between the above examples and their novel presentation in the form of calculi (\emph{coeffect systems}).
The presented type systems and formal semantics highlight the relationship between different
notions of context.

Our second contribution is the definition of two unified coeffect calculi that capture the common
structure of the examples. In particular, our \emph{flat coeffect calculus} models languages
with contextual properties of the execution environment and our \emph{structural coeffect
calculus} models languages where the contextual properties are attached to the variable usage.
We define the semantics of the calculi in terms of category theoretical structure of an
\emph{indexed comonad} (based on dualisation of the well-known monad structure), use it
to define operational semantics and prove safety result for the calculi.

Our third contribution is a novel presentation of our work in the form of web-based
\emph{interactive essay}. This provides a simple implementation of three context-aware programming
languages and lets the reader write and run simple context-aware programs, but also explore the
theory behind the implementation including the typing derivation and semantics.
